---
title: "Practical 17"
author: "Ndivhuwo Nyase"
date: "2023-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("caTools")  
#install.packages('caret')
library(tidyverse)
library(caTools)
library(corrplot)
library(ggplot2)
set.seed(1)
```

## Question 1
### 1) Download the pima data from Vula under Resources –> data and import it into R.

```{r Question 1}
pima <- read.csv("~/Desktop/statistics-computing-R/datasets/pima.csv")
```

## Question 2

```{r q2}

head(pima)
summary(pima)


```

```{r question 2 plot}
pima_corr <- cor(pima,pima)
pima_corr
corrplot(pima_corr,method = 'circle',addCoef.col ='black', number.cex = 0.6)



```
We will take the highly correlated vairables(GLC and age) and plot them against if the subject has diabetes. 
```{r scatterplot }
ggplot(data=pima, mapping = aes(x = glc, y = age)) + 
  geom_point(aes(color = class,group = bmi, size = 2)) +
  ggtitle("Plot showing relationship of GLC and age highlighed by if subject has diabetes.") +
  xlab("Age(years)") + ylab("GLC")
```


```{r barplot}

par(mfcol = c(2,2))
for (index in 1:ncol(pima)){
  
 if (colnames(pima[index]) == "class"){ 
   break 
 }else {  boxplot(formula = unlist(pima[index])  ~ pima$class,
        col = index+1,
        ylab = paste0("",colnames(pima[index])),
        xlab= 'Class',
        main = paste0('Relationship between diabetes and ',colnames(pima[index]) )
 )}
}

```

## Question 3
### Split the data into a training set, a validation set and a test set, with a[60:20:20] ratio.

```{r Question 3}
dtrain <-  sort(sample(nrow(pima), nrow(pima)*.6, replace = FALSE))
train<-pima[dtrain,]

dval <- sort(sample(nrow(pima), nrow(pima)*.2, replace = FALSE))
val <- pima[dval,]
dtest <- sort(sample(nrow(pima), nrow(pima)*.2, replace = FALSE))

test<-pima[dtest,]

```

## Question 4
### Using only one explanatory variable, fit a logistic regression model to the training data, using the validation data to assess the predictive accuracy of the model.

In this logistic regression model we will use the variable glucose (glc) in this model as it has the highest correlation and seems to have a stronger relationship on whether the patient has diabetes on the box plots.
```{r Question 4}
val_q4 <- val %>% dplyr::select(glc, class)
logistic_model <- glm(class ~ glc,
                      data = train,
                      family = 'binomial')

summary(logistic_model)
confint(logistic_model)

predict_ <- predict(logistic_model, 
                       val_q4, type = "response")
predict_ 
   
# Changing probabilities
predict_ <- ifelse(predict_ >0.5, 1, 0)

# Evaluating model accuracy
# using confusion matrix
caret::confusionMatrix(table(unlist(val_q4['class']), predict_))

misclassification <- mean(predict_!=val_q4['class'])
print(paste0('Accuracy on Validation Set: ',1-misclassification))
```

```{r question 5 }


library(doParallel)
detectCores()
cl <- makeCluster(3)
registerDoParallel(cl)

#double-check that you are using > 1 core!
getDoParWorkers()

ptime <- system.time({})
model <- glm(class ~ glc + prg + bp,
                      data = train,
                      family = 'binomial')

colnames(train)
mod_summaries_3 <- list()


for(i in 2:ncol(train)) {     
  predictors_i <- colnames(train)[2:i]# Head of for-loop
 for(j in 2:ncol(train)) {
      predictor_j <- colnames(train)[2:j]
      for (k in 2:ncol(train)) {
        predictor_k <- colnames(train)[2:k]
          mod_summaries_3[[i - 1]] <- summary(     # Store regression model summary in list
            lm(class ~ ., train[ , c("class", predictors_i, predictor_j, predictor_k)]))
      }
 
 }
}
 
mod_summaries_3


train_y <- train['class']
train <- subset(train, select = -c(class))
comb_df <- list()
for (index in 1:ncol(train)) {
comb_df[index]<-lapply(as.data.frame(combn(ncol(train) - 1, 2)), function(idx)
    train[, c(1, idx + 1)])
}

comb_df

```
