---
title: "prac12"
author: "Ndivhuwo Nyase"
date: "2023-02-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

What is probability of getting the different winning symbols?
1.1 Spin the reel once.
1.2 Note the symbol.
1.3 Repeat 1 and 2 above many times (10000).
1.4 Take note of how many times each of the symbols appear.
1.5 Divide by 10000. This is an estimate of the probability.

```{r setup of payout and symbol functions}
# Create the reel with the various probabililites
reel <- 1:6 #the different categories
probs <- c( 7, 5, 3, 6, 9, 20 )/50
spin <- sample(reel, size = 1, prob = probs)

payout_and_symbol <- function(spin) {
  
  
  if (spin==1)
  {
    payout<-2
    symbol<- 'hearts'
  } else if (spin==2)
  {
    payout<-2.5
    symbol <- 'diamonds'
  } else if (spin==3)
  {
    payout<-8
    symbol <- 'spades'
  }else if (spin==4)
  {
    payout<-4
    symbol <- 'clubs'
  } else if (spin==5)
  {
    payout<-2.5
    symbol <- 'joker'
  } else if (spin == 6)
  { payout <- 0 
  symbol <- 'ca'
  }else { 
    payout <- 0
    symbol <- FALSE}
  
  return(c(payout,symbol))
}
```

## QUESTION 1

```{r pressure, echo=TRUE}
# Part 1 
# 1.1 Spin the reel once.
spin <- sample(reel, size = 1, prob = probs)
spin
symbol<- payout_and_symbol(spin)
symbol[2]

spin_reel_nsim <- function(simulations,payout=FALSE,symbol=TRUE){
  i <- 0 
  result <- c()
  while(i < simulations) {
    i <- i + 1
    spin <- sample(reel, size = 1, prob = probs, replace = TRUE)
    p_s <- payout_and_symbol(spin)
    if (symbol == TRUE & payout == FALSE)
    {
      result <- append(result,p_s[2])
    } else if (payout == TRUE & symbol == FALSE) {
      result <- append(result,p_s[1])
    } else if (payout == TRUE & symbol == TRUE) {
      result <- append(result,p_s)
    } else return('CHECK PARAMETERS')
    
  }
  
  return(result)
}

#1.3
part1 <- spin_reel_nsim(10000)


part1<-as.factor(part1)
# 1.4
tab <- table(part1)
tab

# 1.5
nsim<-sum(as.vector(table(part1)))

probabilities <- (tab/nsim)*100
cbind(levels(part1),probabilities)
```
## QUESTION 2
Calculate the average pay out if you spin the reel once. - i.e. one ‘play’ of the
game is to spin the reel once, and you play the game 10000 times.
2.1 Spin the reel once.
2.2 Note the symbol.
2.3 Note the payout – store it.
2.4 Repeat 1, 2 and 3 above many times (10000).
2.5 You should now have 10 000 payout values.
2.6 Calculate the mean of these values. This is an estimate of the average.
payout if you play the game once. - Compare it to the true value!

```{r}
# Part 2
# 2.1
spin <- sample(reel, size = 1, prob = probs)
spin
#2.2 and 2.3
p_s <- payout_and_symbol(spin)
p_s[1]

#2.4
part2 <- spin_reel_nsim(10000, payout = TRUE, symbol = FALSE)

part2f <- as.numeric(part2)

mean(part2f)

#2.5
spin <- sample(reel, size = 10000, prob = probs, replace = TRUE)

true <- mean(probs[spin])
true
```
## QUESTION 3
Calculate the average total pay out if you spin the reel 10/100/1000 times -
i.e. one ‘play’ of the game is to spin the reel 10/100/1000 times, and you play
the game 10000 times.
3.1 Graphically display the average total pay out after 10/100/1000 plays of the
game by means of histograms and box and whisker plots.
3.2 Describe the distribution of the average total pay out pay out for each case
by calculating relevant summary statistics.

```{r}
ngames_func <- function(ngames, size_, payout_=TRUE, symbol_=FALSE){
  game_pay <- numeric(ngames)
  for (i in 1:ngames) {
    outs = as.numeric(spin_reel_nsim(size_,payout = payout_, symbol = symbol_ ))
    game_pay[i] <- mean(outs)
  }
  return(game_pay)
}

game10 <- ngames_func(10000,10,payout_ = TRUE,symbol_ = F)
hist(game10)


game100 <- ngames_func(10000,100,payout_ = TRUE,symbol_ = F)
hist(game100)

game1000 <- ngames_func(10000,1000,payout_ = TRUE,symbol_ = F)
hist(game1000)

boxplot(game10,game100,game1000, frame = FALSE,
        notch = TRUE, names = c('10 spins','100 spins', '1000 spins')
        , col =2:4 )

all_games <- cbind(game10,game100,game1000)
summary(all_games)


```
The distribution of game 10 is more extreme as you can get averages of games from 0 to 4.9. This is due to the fact that in one game repeated 10 times you may get lucky or very unlucky. 

The distribution of game 100 and 1000 are more centred towards the mean due to cental limit theorom in which with a larger sample size, your sample mean is more likely to be close to the real population mean.



