---
title: "Introduction: Statistical Computing with R"
subtitle: "Part  2: matrices and data frames"
author: "Aadapted from Birgit Erni, Allan Clark, Res Altwegg"
institute: "Department of Statistical Sciences, University of Cape Twon "
date:  "`r format(Sys.time(), '%d %B %Y')`"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Outline

- matrices and matrix algebra
- data frames

## Matrices

$$X = \begin{bmatrix} 1 & 1.2 & 3 \\
                      1 & 5.6 & 10 \\
                      1 & 3.5 & 7 \\
      \end{bmatrix}$$

e.g. regression (design) matrix

## Creating Matrices

1. Convert a vector into an n x m matrix. The following code will create a $4 \times 4$ matrix. Note that the matrix is created column wise, i.e. column one is filled up first, etc.. You can change this by specifying `byrow = T`.

```{r matrix1, echo = TRUE}
numbers <- 1:16
m1 <- matrix(numbers, nrow = 4)
```

2. With cbind() or rbind(): binds vectors column- or row-wise
```{r matrix2, echo = TRUE}
numbers1 <- 1:4
numbers2 <- 11:14
m2 <- cbind(numbers1, numbers2)  # column-bind
```

3. diagonal matrix
```{r matrix 3, eval = FALSE, echo = TRUE}
diag(4)          # 4 by 4 identity matrix
diag(c(0.1, 0.2, 0.7))
```

## Calculations with matrices

```{r matrix4, eval = FALSE, echo = TRUE}
t(m1)                # transpose of m1
dim(m1)              # 4 by 4 matrix

m1 * m1          # element-wise multiplication

v1 <- rep(1, times = 4)
v1 %*% m1         # matrix multiplication
m1 %*% m1

m1 <- matrix(1:4,nrow = 2)
solve(m1)          # inverse of m1
solve(m1) %*% m1       

rowSums(m1)
colMeans(m1)
```

## Subsetting/indexing matrices

```{r matrix5, eval = FALSE, echo = TRUE}
m3 <- matrix(1:12, nrow = 3)
m3[2, ]     # row 2
m3[, 3]     # column 3
m3[2, 3]    # element at row 2, column 3
m3[-1, ]     # all except row 1
```

## Prac 2

There are 2 trees in the middle of the Kalahari. On each tree birds of unknown species are sitting and feeling very hot. A bird from the first tree says to those on the second tree: “Hi – if one of you come to our tree then there will be the same number of us on each tree”. “Yeah,
right”, says a bird from the second tree, “but if one of you comes to our tree, then
we will be twice as many on our tree as on yours”.[^1]

**Question:** How many birds are on each tree? More specifically:

- Write up two equations with two unknowns.
- Solve these equations using the methods you have learned above.
- Simply finding the solution by trial–and–error is considered cheating.


[^1]: adapted from <https://www.math.uh.edu/~jmorgan/Math6397/day13/LinearAlgebraR-Handout.pdf>

## Logical operators

!  &  |  == 

```{r logical, echo = TRUE, eval = FALSE}
y <- c(1, 2, 3, NA, 4, 5)

is.na(y)
!is.na(y)

y > 1 & y < 2

y > 1 | y < 2

y[!is.na(y)]

```

## Character vectors

```{r char1, eval = FALSE, echo = TRUE}
site <- c("CT", "JHB", "JHB", "DB")
filename <- "Rintro.R"    # vector of length 1
lett <- letters; letters
treat <- rep(LETTERS[1:3], each = 3); treat
```


## Types of R objects

- **data objects/structures:** vectors, matrices, data frames, lists, arrays, factors
- **functions**
- output from regression model 


## Data frames

- This is the way statisticians, and R, like to store data.
- Like a spreadsheet (rows and columns).
- Ideally: each row is an observation (case), each column is a variable (*tidy data*).
- Similar to matrices, but columns can be of different data types. 

Example (one data set from R, one from Gapminder):

```{r df, echo = TRUE, eval = TRUE}
head(airquality)
```

## Data frames

```{r df2, echo = TRUE, eval = TRUE}
library(gapminder)
str(gapminder)
head(gapminder)
```

## Extracting values from data frames

similar to matrices

```{r df3, echo = TRUE, eval = FALSE}
names(gapminder)
gapminder$country  # only the country column, now a vector
gapminder["country"]

## only SA rows
gapminder[gapminder$country == "South Africa", ]  
```


## Some basic statistics

**Frequency Tables**
```{r basicstats, echo = TRUE, eval = FALSE}
table(continent)
table(gapminder$continent)
with(gapminder, table(continent))

## avoid attach! 
attach(gapminder)
table(continent)
detach(gapminder)
```

## Summary Statistics

```{r basicstats2, echo = TRUE, eval = TRUE}
mean(gapminder$lifeExp)
summary(gapminder$gdpPercap)  # 5-number summary
```

sd, var, min, range, quantile


## Prac 3

\footnotesize
1. Find all rows in airquality that have missing values. 
2. Find mean, sd, min, max for each of temperature and ozone level.
3. For **linear regression**, parameter estimates can be found as follows. 

$$\hat{\beta} = (X'X)^{-1} X'Y$$

Here, $Y$ is the response variable, and $X$ is the design matrix. 

The `cars` data (an R data set) contains two variables: speed and distance to stop. Fit a simple linear regression model to these data, i.e. find the $\hat{\beta}$ estimates, using the equation above, and matrix calcuations in R. 

4. Check that you get the same $\hat{\beta}$ estimates as when fitting the linear regression model using `lm()` in R.

```{r lm, eval = FALSE, echo = TRUE}
m1 <- lm(dist ~ speed, data = cars)
summary(m1)
```


5. Mean life expectancy in South Africa? 53.99317.




